---
title: "Introduction to popler"
author: "Aldo Compagnoni, Sam Levin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to popler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The popler R package is an interface that allows browsing and querying population data collected at Long Term Ecological Research (LTER) network sites located across the United States of America. A subset of the population data from the LTER network is contained in an online database called `popler`. The popler R package is an interface to this online database, allowing users to:

- explore *what* type of population data is contained in the `popler` database
- download data contained in the `popler` database
- filter and validate the data once it is downloaded


## Installation

The popler R package is currently in the development phase, and it should be downloaded directly from its [GitHub page](https://github.com/AldoCompagnoni/popler). Before doing this, make sure to install the [Devtools R package](https://cran.r-project.org/web/packages/devtools/README.html). Once devtools is installed on your machine, install and load popler:

```{r,warnings = FALSE, message = FALSE}
# devtools::install_github("AldoCompagnoni/popler", build_vignettes = TRUE)
library(popler)
```

## Metadata: *what* type of data is contained in the popler database?

The LTER network provides data from thousands of research projects. The metadata of these projects allow understanding *what* population data is provided by each project. The `popler` R package provides three functions to explore these metadata.

### pplr_dictionary()

`pplr_dictionary()` shows:

- what are the variables contained in the metadata of each project and their meaning 
- what data these variables contain.

To see metadata variables and their meaning:  

```{r}
pplr_dictionary()
```

To show what data each variable actually contains, specify one or more variable:

```{r}
pplr_dictionary(lterid, duration_years)
```

Last, but not least, the same information provided by `pplr_dictionary` can be visualized in an html page containing hyperlinks. To open such html page, execute the `pplr_report_dictionary` function. 

```{r eval = FALSE}
pplr_report_dictionary()
```

### pplr_browse()

`pplr_browse()` accesses and subsets the popler metadata table directly. Calling the function returns a table that contains the metadata of all the projects in `popler`:

```{r, eval = FALSE}
all_studies <- pplr_browse()
```

This metadata table can be subset by specifying a logical expression. This is useful to focus on datasets of interest.

```{r}
poa_metadata  <- pplr_browse(genus == "Poa" & species == "fendleriana")
poa_metadata
```

Moreover, akin to `pplr_report_dictionary()`, browse can generate a report and open it as an html page. To do so, set the `report` variable to `TRUE`. Alternatively, you can pass an object created by `browse()` to `pplr_report_metadata()` to create the same report.  

```{r eval = FALSE}

pplr_browse(lterid == "SEV", report = TRUE)

SEV <- pplr_browse(lterid == "SEV")

pplr_report_metadata(SEV)

```

#### The keyword argument

`pplr_browse()` can also single out projects based on partial matching of all of the variables that are made of strings. Specify the string you want to search using the `keyword` argument: 

```{r eval = FALSE}
pplr_browse(keyword == "parasite", report = TRUE)
```

## Download data

Once you identified one or more datasets of interest, download their raw data using `pplr_get_data()`. You can download data in three ways:

1. Passing an object created through `pplr_browse()`

2. Passing an object created by `pplr_browse()` and an additional logical expression to further subset this

3. Passing a logical expression directly to `pplr_get_data()`

A few examples on how to the use of `pplr_get_data()` are:

``` {r eval = FALSE}
poa_metadata    <- pplr_browse(genus == "Poa" & species == "fendleriana") 
poa_data        <- pplr_get_data(poa_metadata) # option 1
poa_data_09     <- pplr_get_data(poa_metadata, year > 2008) # option 2
parasite_data   <- pplr_get_data(proj_metadata_key == 25) # option 3

```

### Carefully vet the methods of downloaded data sets.

We urge the user to carefully read the documentation of each project before using it for research purposes. Data sets downloaded with `popler` share data structure, but each project has its peculiarities. To show the *metadata* of the downloaded data sets, use `pplr_report_metadata` on the data object produced by `pplr_get_data()`. To read the *methods* of each project, click on  the 'metadata link' hyperlink provided in the html page.

```{r eval = FALSE}
pplr_report_metadata(poa_data)

```

### Data structure

All objects produced by `pplr_get_data()` have the same structure. Most variable names are intuitive, but there are at least two nonintuitive variable types.

1. Variable names that match the pattern `spatial_replication_level_` refers to *nested* levels of spatial replication. For example, `spatial_replication_level_1` can represent a site, and `spatial_replication_level_2` represents a plot. In this specific case, `spatial_replication_level_1_label` will contain the string 'site', and `spatial_replication_level_2_label` will contain the string 'plot'.

2. For data sets that contain data on multiple species, data on `genus` and `species` are not always available. In cases when the `genus` and `species` variables are not available, taxonomic identifiers will be contained in the `sppcode` variable. This variable contains codes for species or other higher taxonomic unit. These codes are string of usually a few characters.

Users can explore the level of temporal replication at each nested level of spatial replication using 
the `pplr_site_rep()` and `pplr_site_rep_plot()` functions. `pplr_site_rep` offfers the user control over level of spatial replication and the frequency and duration of temporal replication to either subset a `pplr_get_data()` object or create a summary table. `pplr_site_rep_plot()` shows the user whether a given site (`spatial_replication_level_1`) was sampled in a given year. For additional examples on how to explore and vet `popler` data, see `vignette('vetting-popler', package = 'popler')`


## Extra covariates

Most data sets provided by the USA LTER network contain more data than the standard variables accounted for by the standard schema of `popler`. In order not to loose the original data, `popler` stored all of the extra information in a character variable named `covariates`. The `popler` package provides two ways to format these covariates into a data frame: the `cov_unpack` argument in `pplr_get data()`, and the `pplr_cov_unpack()` function in `popler`.
Setting the `cov_unpack` argument to TRUE returns a data frame that combines the variables of a default query to popler, and the covariates contained in each particular study downloaded through popler.

``` {r}
d_47_cov <- pplr_get_data(proj_metadata_key == 47, cov_unpack = TRUE)
head(d_47_cov)
```

Using the `pplr_cov_unpack()` function on a data frame downloaded using `pplr_get_data()` returns a _separate_ data frame of the covariates contained in the downloaded object.

```{r}
d_47 <- pplr_get_data(proj_metadata_key == 47)
head(pplr_cov_unpack(d_47))
```
