---
title: "Notes"
author: "Aldo Compagnoni"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
## The argument `full_tbl`

Metadata information in popler consists of 75 variables, but only 22 of these are shown by the default settings of the package. The 22 "default" variables are the most intuitive and the most commonly used in database searches. However, if you want to give a deeper look into project metadata, you only need to set `full_tbl = TRUE` in functions `dictionary()` and `browse()`:

```{r, warning = FALSE, message = FALSE}
library(popler)
pplr_dictionary(full_tbl = TRUE)
pplr_browse(full_tbl = TRUE)
```

## What does `full_tbl = FALSE` leave out

Controlling which variables are "left out" by setting `full_tbl = FALSE` reveals a few patterns: 

```{r, warning = FALSE, message = FALSE}
left_out <- setdiff(pplr_dictionary(full_tbl = TRUE)$variable, pplr_dictionary()$variable)
keep     <- which( pplr_dictionary(full_tbl = TRUE)$variable %in% left_out )
pplr_dictionary(full_tbl = TRUE)[keep,]
```

The type of variables left out can be lumped into four categories: spatio-temporal replication, measurement units, project specific information (PI, contacts, information on LTER site), and taxonomic information. In each of these cases, the variables left out are unlikely to be used in searches. For example, the variable "superdivision" is a taxonomic group provided by only a very small subset of studies, and it is seldom used in scientific publications (providing "superdivision" into ISI web of Science yields only two scientific articles).

## The `schema()` function

`schema()` returns a description of all the variables contained in popler. It returns a list where each element refers to a specific table of the data base. Each element is a data frame with two columns: the names of the variables contained in the table, and its description.
Differences in both 

- `schema()` can also open a figure of the graphical representation of the schema of popler.
- Or should we just save the schema as data using devtools::use_data?

