---
title: "Vetting popler"
author: "Aldo Compagnoni, Sam Levin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vetting popler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction: identifying groups of data sets

The `popler` R package was built to foster scientific synthesis using LTER long-term population data. The premise of such synthesis is using data from many research projects that share characteristics of scientific interest. To identify projects sharing salient attributes, `popler` uses the metadata information associated with each LTER project. In particular, it is farily easy to select projects based on one or more of the following features:

 1. Replication, temporal or spatial.
 2. Taxonomic group(s).
 3. Study characteristics. 
 4. Geographic location.

Vetting the database based on these criteria is intuitive. However, `popler` also facilitates identifying data sets in other ways. Below we provide several examples on how to select LTER data based on the four types of features described above. Moreover, in the final section we also show how to carry out more complicated types of searches.

### 1. Replication
#### Temporal replication

If you are interested in long-term data, you will likely want to select projects based on how many years the data was collected for. This is straightforward: 

```{r, warning = FALSE, message = FALSE}

library(popler)
pplr_browse(duration_years > 10)

```

Note that most LTER projects contemplate sampling at a yearly or sub-yearly frequency. Thus, studies longer than 10 years generally guarantee a longitudinal series of 10 or more observations. Note that the `duration_years` variable is calculated as `studyendyr - studystartyr`. Thus, an additional variable named `samplefreq` characterizes the approximate sample frequency of each study. 

```{r, warning = FALSE, message = FALSE}

pplr_dictionary(samplefreq)
pplr_browse(samplefreq == "monthly")

```

Note that `samplefreq` is **not** a default variable included in the `pplr_dictionary` or `pplr_browse()` functions. This can be viewed by specifying the `full_tbl = TRUE` argument in either function.

###1. Spatial replication

#### Before downloading data

If you wish to select data sets based on their spatial replication, you need to consider that `popler` organizes data in nested spatial levels. For example, in many plant studies data is collected at the *plot* level, which can be nested within *block*, which in turn can be nested within *site*. `popler` labels spatial levels using numbers: 1 is the coarsest level of replication which contains all other spatial replicates. In the example above, spatial level 1 is *site*, spatial level 2 is *block*, and spatial level 3 is *plot*. `popler` allows for a total of 5 spatial levels. Given the above, you can select studies based on three criteria:

 1. The total number of spatial replicates.
 
 2. The number of replicates within a specific spatial level.
 
 3. The number of nested spatial replicates.

Below we provide respective three examples for each one of these respective searches.

```{r}
pplr_browse(tot_spat_rep > 100)
pplr_browse(spatial_replication_level_5_number_of_unique_reps > 1)
pplr_browse(n_spat_levs == 3)
```

#### After downloading data

Users can also explore the spatial and temporal replication of the data more explicitly after downloading it with `pplr_get_data()`. `pplr_site_rep()` provides two options for exploring data that meet temporal replication requirements at a given spatial resolution. You can choose to filter your data by specifying a minimum sampling frequency per year and a minimum number of years that sample with that frequency. Because this function uses the sampling dates to calculate the frequency, it provides additional information that is not contained in the `samplefreq` column of the main metadata table.


```{r, eval = FALSE}
# download some data (note: this download is >100MB)
SEV <- pplr_get_data(proj_metadata_key == 21)

# Create a summary table containing names of replication levels that contain 2 samples per year for 10 years. 
SEV_long_studies <- pplr_site_rep(SEV, 
                                  freq = 2, 
                                  duration = 10, 
                                  return_logical = FALSE)

# you can also subset it directly using the function and specifying it to return a logical vector
subset_vec <- pplr_site_rep(SEV,
                            freq = 2,
                            duration = 10,
                            return_logical = TRUE)
# store subset of data
SEV_long_data <- SEV[subset_vec, ]



```


Users can also visualize the frequency of sampling at the coarsest level of spatial replication using the `pplr_site_rep_plot()` function. This generates a `ggplot` that denotes whether or not a particular site was sampled in a particular year.

```{r, eval = FALSE}

library(ggplot2)

# return the plot object w/ return_plot = TRUE
pplr_site_rep_plot(SEV_long_data, return_plot = TRUE) +
  ggtitle("Long Term Data from Sevilleta LTER")
  
# or return an invisible copy of the input data and keep piping
library(dplyr)
SEV_long_data %>%
  pplr_site_rep_plot(return_plot = FALSE) %>%
  pplr_report_metadata()

```


###2. Taxonomic group

`popler` is not limited to specific taxonomic groups, and it currently contains mostly data on animals and plants. To select information based on taxonomic groups, simply specify which group and which category you wish to select. The default settings of popler provide seven taxonomic groups: kingdom, phylum, class, order, family, genus, and species in each request. Column `sppcode` provides the identifier, usually an alphanumeric code, associated with each taxonomic entity in the original dataset.
Note that not all LTER studies provide full taxonomic information; hence, browsing studies by taxonomic information will provide partial results (in the example below, not all insects studies will be identified).

```{r, warning = FALSE, message = FALSE}
pplr_dictionary(class)
pplr_browse(class == "Insecta")
```

Note that taxonomic information returned in `pplr_browse()` is housed in a data structure called a _list column_. Each entry of this list column is itself a list that contains a `data.frame` with eight columns. Users can access this information using the following syntax.

```{r, warning = FALSE, message = FALSE}

insects <- pplr_browse(class == 'Insecta')

# access the taxonomic table from the first project in the insects object
insects$taxas[[1]]

# second table
insects$taxas[[2]]

```


###3. Study characteristics

Metadata information provides a few variables to select studies based on their a straightforward way. In particular:

 - `studytype`: indicates whether the study is observational or experimental. 
 - `treatment_type`: type of treatments, if study is experimental.
 - `community`: does project provide data on multiple species?

Below we show
how to use these three fields. 

```{r, warning = FALSE, message = FALSE}
pplr_dictionary(community)
pplr_browse(community == "no") # 20 single-species studies

pplr_dictionary(treatment)
nrow( pplr_browse(treatment == "fire") ) # 21 fire studies

pplr_dictionary(studytype)
nrow( pplr_browse(studytype == "obs") ) # 78 observational studies
```

### 4. Geographic location.

To select studies based on the latitude and longitude of the LTER headquarters, simply use the `lat_lter` and `lng_lter` numeric variables: 

```{r, warning = FALSE, message = FALSE}
pplr_dictionary( lat_lter,lng_lter )
pplr_browse(lat_lter > 40 & lng_lter < -100) # single-species studies
```

### 5. More complicated searches

Popler allows carrying out more complicated searches by allowing to i)  simultaneously search several types of metadata variables, and ii) search studies matching a string pattern. In the first case, simply provide the function `pplr_browse()` with a logical statement regarding more than one metadata variable. For example, if you want studies on plants with at least 4 nested spatial levels, and 10 years of data:

```{r, eval = FALSE}

pplr_browse(kingdom == "Plantae" & n_spat_levs == 4 & duration_years > 10)

```

In the second case, the keyword argument in function `pplr_browse()` will search for string patterns within the metadata of each study. For example, in case we were interested in studies using traps:

```{r, eval = FALSE}

pplr_browse(keyword = 'trap')

```

Note that the keyword argument works with regular expressions as well:

```{r, eval = FALSE}

# look for studies that include the words "trap" or "spatial"
pplr_browse(keyword = 'trap|spatial')

```
