---
title: "Vetting popler"
author: "Aldo Compagnoni"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction: identifying groups of data sets

The popler R package was built to foster scientific synthesis using LTER long-term population data. The premise of such synthesis is using data from many research projects that share characteristics of scientific interest. To identify projects sharing salient attributes, popler uses the metadata information associated with each LTER project. In particular, it is farily easy to select projects based on one or more of the following features:

 1. Replication, temporal or spatial.
 2. Taxonomic group.
 3. Study characteristics. 
 4. Geographic location.

Vetting the database based on these criteria is very intuitive. However, popler also facilitates identifying data sets in many other, more flexible ways. Below we provide several examples on how to select LTER data based on the four types of features described above. Moreover, in the final section we also show how to carry out more complicated types of searches.

### 1. Replication
#### Temporal replication

If you are interested in long-term data, you will likely want to select projects based on how long the data was collected for. This is very straightforward using the 

```{r, warning = FALSE, message = FALSE}
library(popler)
browse(duration_years > 10)
```

Note that most LTER projects contemplate sampling at a yearly or sub-yearly frequency. Thus, studies longer than 10 years generally guarantee a longitudinal series of 10 or more observations. The variable "samplefreq" characterizes the sample frequency of each study. [COMMENT: "samplefreq" is not browse() and dictionary() default settings. Still debating whether I should describe *full_tbl* right away or below]: 

```{r, warning = FALSE, message = FALSE}
dictionary("samplefreq")
browse(samplefreq == "monthly")
```

#### Spatial replication

If you wish to select data sets based on their spatial replication, you need to consider that popler organizes data in nested spatial levels. For example, in many plant studies data is collected at the *plot* level, which can be nested within *block*, which in turn can be nested within *site*. Plot labels spatial levels using numbers: 1 is the highest level of replication which contains all other spatial replicates. In the example above, spatial level 1 is *site*, spatial level 2 is *block*, and spatial level 3 is *plot*. Popler allows for a total of 5 spatial levels. Given the above, you can select studies based on three criteria:

 1. The total number of spatial replicates.
 1. The number of replicates within a specific spatial level.
 1. The number of nested spatial replicates.

Below we provide respective three examples for each one of these respective searches:

```
browse(tot_spat_rep > 100)
browse(spatial_replication_level_1_number_of_unique_reps > 5)
browse(n_spat_levs == 3)
```

###2. Taxonomic group

Popler contains a wealth of taxonomic information......

To select information based on taxonomic groups, simply specify which group and which category you wish to select. The default settings of popler provide seven taxonomic groups: kingdom, phylum, class, order, family, genus, and species: 

```{r, warning = FALSE, message = FALSE}
dictionary("class")
browse(class == "Insecta")
```

###3. Study characteristics

Metadata information provides a few variables to select studies based on their a straightforward way. In particular:

 - `studytype`: indicates whether the study is observational or experimental. 
 - `treatment_type`: type of treatments, if study is experimental.
 - `community`: does project provide data on multiple species?

Below we provie several three examples on how to use these three fields. 

```{r, warning = FALSE, message = FALSE}
dictionary("community")
browse(community == "no") # 8 single-species studies

dictionary("treatment_type_1")
nrow( browse(treatment_type_1 == "grazing") ) # 2 grazing studies

dictionary("studytype")
nrow( browse(studytype == "obs") ) # 28 observational studies
```

### 4. Geographic location.

To select studies based on the latitude and longitude of the LTER headquarters, simply use the `lat_lter` and `lng_lter` numeric variables: 

```{r, warning = FALSE, message = FALSE}
dictionary( c("lat_lter","lng_lter") )
browse(lat_lter > 40 & lng_lter < -100) # single-species studies
```

### 5. More complicated searches

Popler allows carrying out more complicated searches than those suggested above. In particular, it allows to i) combine to simultaneously search several types of metadata variables, and ii) use keywords. In the first case, simply provide the function `browse()` with a logical statement regarding more than one metadata variable. For example, if you want studies on plants with at least 4 nested spatial levels, and 10 years of data:

```{r, warning = FALSE, message = FALSE}
browse(kingdom == "Plantae" & n_spat_levs == 4 & duration_years > 10)
```

In the second case, using keywords can be crucial to retrieve data types that are not directly related to the information contained in metadata variables. For example, you might want to select data based on a particular research topic, rather than the taxa or the experimental design. To aid these searches, popler provides the `keyword` binary operator `%=%`. tHIS O 

```
browse( title %=% "parasit" & n_spat_levs > 2 | genus == "Eimeria")
```
